Supplementary Materials for \"COATi: statistical pairwise alignment of protein coding sequences\

### 2024-02-09 response_r1.md

---
title: Response to Reviewers
subtitle: "COATi: statistical pairwise alignment of protein-coding sequences"
author:
	- Juan J. García Mesa
	- Ziqi Zhu
	- Reed A. Cartwright
mainfont: TeX Gyre Pagella
geometry:
	- margin=1in
---

<!-- pandoc --pdf-engine=lualatex -o response_r1.pdf response_r1.md -->

Dear Editor,

We are resubmitting our manuscript, "COATi: statistical pairwise alignment of protein coding sequences", for your consideration. This manuscript describes a new model for aligning protein coding sequences which produces more accurate alignments, while being robust to common artifacts found in genomic annotations. Our approach is implemented in the software package, COATi, which is released via an MIT license and available on GitHub. While COATi can generate multiple-sequence alignments, this paper is focused on our pairwise aligner which is has received more attention and validation. Additionally, to promote open science and reproducibility all data, procedures, and scripts used to validate COATi and compare it with alternative methods have been uploaded to GitHub.

We would like to thank the associate editor and two reviewers for their helpful comments and suggestions. As requested, we have broadened the scope of the manuscript from a software note to a larger research paper about our COATi software. Nearly all of the paper has been rewritten, expanded, and reconfigured to provide more details and background about our work. Importantly, we now use empirical results from the human-gorilla sequence pairs to initially compare aligners before we show results on semi-empirical benchmarks. If certain concepts need additional exposition, we will be happy to make those adjustments.

We have addressed all of the concerns brought to our attention by the associate editor and reviewers as described below.

Sincerely,  
Reed A. Cartwright, on behalf of all authors

## Associate Editor Comments

@@

 1. Comment: *The general consensus about this paper is that the question and method (codon-aware alignment) are both extremely important and suitable for MBE, but the manuscript, written essentially as a short software note, is simply too sparse to adequately present the solution, eliding over important issues and methodological details.*

    We thank the editor and reviews for the positive feedback. As mentioned above, we have now broadened the scope of the manuscript from a software note to a larger research paper about our COATi software. We have rewritten nearly the entire paper as we expanded to scope in response to feedback.

 1. Comment: *In addition to the reviewers' comments, one more critical question strikes me. The method makes uses of a simplification procedure predicated on the concept of treating one sequence as if it is the ancestor of the other, which allows modeling the problem using FSTs rather than HMMs. This modeling choice introduces additional directionality into the model. If there are two sequences (A and B) the method arbitrarily chooses one for the ancestor and one for the descendant. What is not at all clear is whether this directionality is important, i.e., is the same result generated if the model goes A-->B as it would if it went B-->A. This is a fundamentally critical question that needs to be addressed as the stability of any alignment method should not be dependent on the order of the sequences within the input file.*

    The editor is correct that our model is directional. This is not a limitation of the model but rather a feature. We have added more details about our FSTs and our reasoning for using them. In particular, we now explain that our input sequence (i.e. "the ancestor" or "reference") is required to have a length that is a multiple of 3 and missing stop codons and ambiguous nucleotides (circa line 99). The output sequence (i.e. "the descendant" or "non-reference") does not have these restrictions. This asymmetry is due to COATi needing one high-quality reference sequence to establish reading-frame contexts.

    In order to test the impact of directionality on our results, we have included results when using gorilla as the reference sequence. The results are very similar to using human as the reference sequence.

@@

## Reviewer 1 Comments

 1. Comment: *It is surprising this work has not been extended to something larger*

    We appreciate the reviewer's suggestion that our ideas may be suitable for longer sequences. We feel that COATi, with a bit of work, would be the correct tool for aligning long-read sequences generated from target genes against a reference. Our existing suite of FSTs can support both exonic and intronic regions, and via concatenation we can create a new FST that supports specific reference genes' exon and intron segments. The challenge is to figure out the best user-interface for specifying such information as well as validating the result. We choose to tackle these challenges in a future research project.

    We now comment on this in the discussion.

## Reviewer 2 Comments

@@

1. Comment: *I would like to start this review with an apology. This review took me a bit longer than I had hoped.*

    We also apologize for taking longer than expected to submit this revision, as the first author needed to focus on defending his dissertation last semester. We are happy to report that he passed his defense without issue and is now Dr. García Mesa, joining Dr. Zhu who graduated the year before.

    We also discovered a bug in COATi's scoring algorithm (used for defining "perfect alignments" in the analysis) as well as minor issues with how we were using MACSE and ClustalΩ. We opted to take the time to regenerate our results after fixing these issues. Tables and figures barely changed and no conclusions were affected.

@@

 1. Comment: *I am confused by the state space. Perhaps they could be enumerated in the supplementary information. Obviously, there are 61 possible codon states (assuming the standard code). It strikes me that there are 12 possible 1 nucleotide, 2 gap states (i.e., N--, -N-, and --N, where N is any nucleotide) and 48 possible 2 nucleotide, 1 gap states (i.e., NN-, N-N, and -NN). I would assume that -N- and N-N would be very rare, but I am fine with including them for the purpose of completeness. There is also the all-gap state, which brings the total up to 122 states.*

    Our FSTs do not have the same state space concept as in phylogenetic models. Consider the codon substitution FST (Fig. 3a). Its input and output sequences have state spaces of four symbols (A, C, G, and T). This FST has two internal states (M and S) and 61x61 arcs from S to M. Each of these 3721 arcs consume three nucleotides from the input sequence, emit three nucleotides to the output sequence, and represent one transition in a standard codon substitution model.

    Similarly, consider the ambiguity FST (Fig 2b). Its input sequence has four symbols (A, C, G, and T) and its output sequence has five symbols (A, C, G, T, and N). It has two internal states (M and S), and 8 arcs from S to M. Four of these arc consume a nucleotide and emit the same nucleotide. The other four arcs consume a nucleotide and emit N. This is the FST that allows COATi to support ambiguous nucleotides in the non-reference/output sequence.

    To handle indels, the indel FST (Fig 3b) has a state space of four symbols (A, C, G, and T) with a special symbol, epsilon, that represents an empty match. An arc that reads nothing (epsilon) and outputs a symbol represents an insertion, while an arc that reads a symbol and outputs nothing (epsilon) represents a deletion.

    We don't specifically model these processes in a single FST as we use the composition algorithm to combine separate FSTs that represent codon substitution, indels, error, and sequencing ambiguity to produce one FST that handles all four processes (Fig 3c).

@@

 1. Comment: *I can see how one can forbid ungapped emission of stop codons. However, I could see problematic cases like …TACAAG… (…Tyr-Lys…) where a deletion of three bases after the first T -or- after the first A would yield a stop codon. Is that forbidden? Or is it allowed in the “low quality sequence” in the pair?*

    Our Indel FST does not consider codon translations when deleting or inserting nucleotides. Indels can occur that create early stop codons. However, this will only happen if the output sequence is of low quality and contains an early stop codon. We consider this a feature of COATi as it will not fail to align output sequences that contain early stop codons.

 1. Comment: *I don’t see how they can use a codon model when indels can occur within codons. This is especially true for the Kosiol ECM. Since the ECM is purely empirical it can be viewed as a matrix of rates of change between 61 symbols. I think this should also be true for the Muse and Gaut codon model.*

    We are able to easily mix codon models with indels that occur within codons because we are modeling alignments using FSTs. The composition algorithm allows us to combine an FST representing a codon substitution model with an FST representing an indel model (Figure 3).  Composition does produce the complex graphs that you would expect from combining a codon model with an indel model, and we show a short example in Figure S1. Composition is a complex but well established algorithm in the FST literature, and we have chosen to refer readers to Mohri et al. (2005) if they would like to know more.

@@

 1. Comment: *It might be possible to improve Fig. 2 to clarify this. It is very difficult to work through, and I am still having trouble despite staring at the figure quite a bit. I wonder if showing the example alignment for Fig. 1 might help. Perhaps something could be done in the supplement.*

    We have updated Figure 2 and added a new Figure 3 that we feel more adequately explains the finite-state transducers that we are using and the composition algorithm that allows us to combine them in a reasonable manner. We also include in the supplement Figures S1 and S2 which show examples of fully composed alignment graph between the codons "CTC" and "CTG" as well as the best path through the graph.

@@

 1. Comment: *Finally, how is the “low quality sequence” in the pair identified?*

    By default, COATi uses the first sequence in the input file as the input/reference sequence and the second sequence as the output/non-reference sequence. It also contains flags to customize this behavior. This gives users the power to place low-quality sequences as the output sequence when aligning a pair of sequences using COATi FST. We now mention this in the manuscript circa line 108.

@@

 1. Comment: *Perhaps one of the biggest problems with the methods is that the first two paragraphs in the results cover a lot of methods. However, they cover them in an abbreviated format that is almost impossible to follow. For example, I think that the way indels are simulated are by taking alignments of empirical human-gorilla sequence pairs and then mapping the gaps from those pairs onto simulated sequences, where those sequences were simulated without gaps. Is that correct? I think the actual methods need to be spelled out much more carefully and put in the methods section.*

    We have added details about how we created our semi-empirical benchmark alignments (circa line 227). The revision now contains information about our procedure, as well as an example of it being applied to data. We believe that this update answers the concerns of the reviewer.

@@

 1. Comment: *In this context, it would be good to state how Ka and Ks are calculated in the methods. The supplement has one sentence about how they are calculated and a lot about what Ka/Ks > 1 means. The material about interpretation is not a problem, but I think many MBE readers will not need it. I’d really like to see an explicit articulation of how the stats were collected in the main paper.*

    We used sequinr to calculate Ka and Ks and have expanded our methods to explain which method that sequinr uses to calculate these stats and how that method works (briefly).

@@

    Back to your question, `coati alignpair` breaks ties according to the implementation of the "shortest path" algorithm in the OpenFST library (when using an FST-based model like tri-mg). We currently use OpenFST to implement our FST-based algorithms. We now mention this in the manuscript (circa line 116).

    Furthermore, `coati sample` doesn't need to break ties as it randomly samples alignments from the posterior. This is useful for studies that need a better picture of alignment space than retrieving a single, best alignment. We've utilized it to study indels across the tree of life (e.g. Zhu 2022), and it will be described in greater detail in a future paper.

 1. Comment: *Finally, I would like to see the impact of alternative alignments on distances – perhaps logdet or K2P or F84 distances. It might be informative to ask how often each alignment method results in the largest distance. It strikes me that methods that “overalign” (e.g., their Fig. 1c) would tend to have the largest distances.*

    *It would also be nice to summarize Ka/Ks and evolutionary distances for empirical human-gorilla alignments. It strikes me that they already have the data, having used it to get parameters for simulation.*

    We would like to thank the reviewer for the suggestion to analyze the empirical human-gorilla alignments. Using this suggestion, we have expanded our comparison of COATi and the other aligners. We now use both empirical human-gorilla sequence pairs and semi-empirical benchmarks. From empirical data, we can confirm that aligners that overalign sequences (fewer gaps, more mismatches) do produce higher K2P distances (Table 1 and Figure 4). For the benchmarks, aligners that had higher alignment errors had higher root-mean-squared errors for K2P distances and lower F1 scores for selection as well.

### 2024-02-09 alignpair_letter.tex
Revision 1 - Release Candidate 1

COATi aligns pairs of sequences using a statistical alignment model, which is implemented as a finite-state transducer derived from the composition of multiple FSTs, each representing a specific biological or technical process: (1) the codon substitution FST, (2) the indel FST, (3) the error FST, and (4) the ambiguity FST \citep[Figs.~\ref{fig:base-calling}--\ref{fig:coati-fst}; c.f.][]{holmes2001evolutionary}. We call this transducer the COATi FST. Codon substitution models are uncommon in sequence aligners, despite their extensive use in phylogenetics. COATi implements the Muse and Gaut (\citeyear{muse_gaut_1994}) codon model (codon-triplet-mg) and the Empirical Codon Model \citep[codon-triplet-ecm;][]{kosiol_ECM_2007}.
It also lets the user provide a codon substitution matrix. A key innovation of COATi is that it combines a codon substitution model with a nucleotide-based indel model, allowing gaps to occur both between and within codons \citep[c.f.][]{ranwez_macse_2011,ranwez_macse_2018}. This also allows the aligner to be robust against sequencing artifacts that produce sequences with disrupted reading frames.

COATi FST is not a true probabilistic FST and cannot be used as-is to simulate output sequences based on an input sequence. This is because it is missing a parameter to control how often Ns are added to the output sequence. This is design feature to allow COATi FST to properly weight ambiguous nucleotides as representing any other symbol. In addition, COATi's indel FST (Fig.~\ref{fig:coati-fst}b) implicitly assumes that match states occur both immediately before and immediately after the alignment. This allows COATi FST to assign the same weight to gaps at the beginning and end of an alignment but also introduces a normalization constant (not shown) to reflect that mass is lost by not allowing the indel FST to terminate from node C. While this normalization constant is not needed to find the most likely alignment or to sample alignments, it can be calculated using a Markov model that considers only the transitions between states M and D in the indel FST (Fig.~\ref{fig:coati-fst}b) and ignoring any states in between:
\[
\frac{1-e + g(e-g)^{n+1}}{1-e+g} \approx \frac{1-e}{1-e+g} 
\]
where $g$ is the gap opening parameter, $e$ is the gap-extension parameter, and $n$ is the length of the input sequence. The exponent has a $+1$ due to the presence of an ``immortal match'' immediately after the input sequence.

Since codon substitution is the first process in COATi's model, the input sequence to the COATi FST must be compatible with a codon-substitution model and be a multiple of three nucleotides and not contain any ambiguous symbols or stop codons. The phases, reading frames, and amino-acid contexts of alignment columns is determined by the input sequence, and better alignments will be generated if the input is of high quality and free of artifacts. Depending on context, we may refer to the input sequence as the ``ancestral'' or ``reference'' sequence. In contrast, the output sequence must be compatible with the ambiguity FST, can be of any length, and can contain any nucleotides or ``N''. This allows COATi to align lower-quality sequences that may contain artifacts against a high-quality reference sequence. We may refer to the output sequence as the ``descendant'', ``non-reference'', or ``query'' sequence. The choice of which sequence is the input sequence and which sequence is the output sequence is left up to the user.

@@

A marginal substitution model is calculated from a standard substitution model by calculating the marginal probabilities that each ancestral codon produces specific descendant nucleotides at each reading frame position. Specifically, let
%
\[
P_\text{cod}\left( Y_1 = y_1, Y_2 = y_2, Y_3 = y_3 |
                   X_1 = x_1, X_2 = x_2, X_3 = x_3 \right)
\]
%
represent transition probabilities from a codon model, and
%
\[
P_\text{mar}\left(Y_p = y \middle| X_1 = x_1, X_2 = x_2, X_3 = x_3 \right)
=
\sum_{y_1, y_2, y_3} I(y_p = y)
P_\text{cod}\left( y_1, y_2, y_3 |
                   x_1, x_2, x_3 \right)
\]
%
represent the marginal transition probabilities, where $p \in \{1, 2, 3\}$ is the position of the descendant nucleotide relative to the ancestral reading frame and $I()$ is an indicator function. COATi contains marginal models for both Muse and Gaut (\citeyear{muse_gaut_1994}) or the Empirical Codon Model, resulting in the marginal models codon-marginal-mg and codon-marginal-ecm. These models emphasize the position in a codon where the substitution occurs, help restrict the effects of low-quality data in the descendant sequence, and allow more than one substitution per codon. In combination with the indel model, alignment using the marginal model is implemented using dynamic programming.

@@

In order to compare COATi against other aligners, we evaluated five different alignment models: (1) COATi's FST model (i.e.\ codon-triplet-mg), (2) the amino-acid model of ClustalΩ via translation and reverse translation, (3) the amino-acid aware nucleotide model of MACSE, (4) the DNA model of MAFFT, and (5) the codon model of PRANK \citep{clustal_omega_sievers_2011,ranwez_macse_2018,katoh2013mafft,prank_loytynoja_2014}. 
COATi is not a symmetric pairwise aligner, as reference sequences are more constrained than non-reference sequences. In order to evaluate the importance of the choice for reference sequence, we also aligned sequences pairs using (6) COATi with gorilla sequences as the reference (i.e.\ COATI-rev). Together, these six methods allow us to evaluate both different alignment strategies and different software implementations. See Supplemental Materials for additional details, including results for COATi's marginal and ECM models.

@@

The distribution of evolutionary distances for COATi, COATi-rev, and PRANK were biologically reasonable  for human-gorilla gene pairs, with means 0.8\%--0.9\% (Fig.~\ref{fig:k2p-empirical}a). Conversely, the means for ClustalΩ, MACSE, and MAFFT were much higher, 1.5\%--1.8\%. These larger means are due to the fact that the right tails of the distributions for ClustalΩ, MACSE, and MAFFT were larger than the other methods. The differences between the distances inferred by COATi and other methods also shows ClustalΩ, MACSE, and MAFFT producing higher distance estimates for a subset of sequences (Fig.~\ref{fig:k2p-empirical}b).
%
COATi's alignments produced significantly lower K2P distances than ClustalΩ, MACSE, MAFFT, and PRANK. Single-tailed Wilcoxon signed-rank tests on matched data produced p-values $< 2.2\times10^{-16}$. Additionally, COATi's K2P distances did not depend on whether human or gorilla was used as the reference sequence (p-value of $0.4212$ for a two-sided test). Additional analysis of the empirical alignments can be found in Supplementary Methods.

@@

In additional to supporting all three phases of gaps, COATi's other advantage over competing aligners in our tests is at least partially due to its default parameters fitting the human-gorilla coding sequences better than the other aligner's default settings. Different data sets may show different results. However, because COATi uses a statistical model with biologically meaningful parameters, users can easily optimize COATi by using parameters estimated by previous studies or estimating parameters themselves \citep[e.g.][]{zhu2022profiling}.

>>>>

COATi aligns pairs of sequences using a statistical alignment model, which is implemented as a finite-state transducer derived from the composition of multiple FSTs, each representing a specific biological or technical process: (1) the codon substitution FST, (2) the indel FST, (3) the error FST, and (4) the ambiguity FST \citep[Figs.~\ref{fig:base-calling}--\ref{fig:coati-fst}; c.f.][]{holmes2001evolutionary}. We call this transducer the COATi FST. Codon substitution models are uncommon in sequence aligners, despite their extensive use in phylogenetics. COATi implements both the \cite{muse_gaut_1994} codon model (codon-triplet-mg) and the Empirical Codon Model \citep[codon-triplet-ecm;][]{kosiol_ECM_2007}.
It also lets the user provide a codon substitution matrix. A key innovation of COATi is that it combines a codon substitution model with a nucleotide-based indel model, allowing gaps to occur both between and within codons \citep[c.f.][]{ranwez_macse_2011,ranwez_macse_2018}. This also allows the aligner to be robust against sequencing artifacts that produce sequences with disrupted reading frames.

COATi FST is not a true probabilistic FST and cannot be used as-is to simulate output sequences based on an input sequence. This is because it is missing a parameter to control how often Ns are added to the output sequence. This is a design feature to allow COATi FST to properly weight ambiguous nucleotides as representing any other symbol. In addition, COATi's indel FST (Fig.~\ref{fig:coati-fst}b) implicitly assumes that match states occur both immediately before and immediately after the alignment. This allows COATi FST to assign the same weight to gaps at the beginning and end of an alignment but also introduces a normalization constant (not shown) to reflect that mass is lost by not allowing the indel FST to terminate from node C. While this normalization constant is not needed to find the most likely alignment or to sample alignments, it can be calculated using a Markov model that considers only the transitions between states M and D in the indel FST (Fig.~\ref{fig:coati-fst}b) and ignoring any states in between:
\[
\frac{1-e + g(e-g)^{n+1}}{1-e+g} \approx \frac{1-e}{1-e+g} 
\]
where $g$ is the gap opening parameter, $e$ is the gap-extension parameter, and $n$ is the length of the input sequence. The exponent has a $+1$ due to the presence of an ``immortal match'' immediately after the input sequence.

Since codon substitution is the first process in COATi's model, the input sequence to the COATi FST must be compatible with a codon-substitution model and be a multiple of three nucleotides and not contain any ambiguous symbols or stop codons. The phases, reading frames, and amino-acid contexts of alignment columns is determined by the input sequence, and better alignments will be generated if the input is of high quality and free of artifacts. Depending on context, we may refer to the input sequence as the ``ancestral'' or ``reference'' sequence. In contrast, the output sequence must be compatible with the ambiguity FST, can be of any length, and can contain any nucleotides or ``N''. This allows COATi to align lower-quality sequences that may contain artifacts against a high-quality reference sequence. We may refer to the output sequence as the ``descendant'' or ``non-reference'' sequence. The choice of which sequence is the input sequence and which sequence is the output sequence is left up to the user.

@@

A marginal substitution model is calculated from a standard substitution model by calculating the marginal probabilities that each ancestral codon produces specific descendant nucleotides at each reading frame position. Specifically, let
%
\[
P_\text{cod}\left( Y_1 = y_1, Y_2 = y_2, Y_3 = y_3 |
                   X_1 = x_1, X_2 = x_2, X_3 = x_3 \right)
\]
%
represent transition probabilities from a codon model, and
%
\[
P_\text{mar}\left(Y_p = y \middle| X_1 = x_1, X_2 = x_2, X_3 = x_3 \right)
=
\sum_{y_1, y_2, y_3} I(y_p = y)
P_\text{cod}\left( y_1, y_2, y_3 |
                   x_1, x_2, x_3 \right)
\]
%
represent the marginal transition probabilities, where $p \in \{1, 2, 3\}$ is the position of the descendant nucleotide relative to the ancestral reading frame and $I()$ is an indicator function. COATi contains marginal models for both \cite{muse_gaut_1994} and the Empirical Codon Model \citep{kosiol_ECM_2007}, resulting in the marginal models codon-marginal-mg and codon-marginal-ecm. These models emphasize the position in a codon where the substitution occurs, help restrict the effects of low-quality data in the descendant sequence, and allow more than one substitution per codon. In combination with the indel model, alignment using the marginal model is implemented using dynamic programming.

@@

In order to compare COATi against other aligners, we evaluated five different alignment models: (1) COATi's FST model (i.e.\ codon-triplet-mg), (2) the amino-acid model of ClustalΩ \citep{clustal_omega_sievers_2011} via translation and reverse translation, (3) the amino-acid aware nucleotide model of MACSE \citep{ranwez_macse_2018}, (4) the DNA model of MAFFT \citep{katoh2013mafft}, and (5) the codon model of PRANK \citep{prank_loytynoja_2014}. 
COATi is not a symmetric pairwise aligner, as reference sequences are more constrained than non-reference sequences. In order to evaluate the importance of the choice for reference sequence, we also aligned sequences pairs using (6) COATi with gorilla sequences as the reference (i.e.\ COATi-rev). Together, these six methods allow us to evaluate both different alignment strategies and different software implementations. See Supplemental Materials for additional details, including results for COATi's marginal and ECM models.

@@

The distribution of evolutionary distances for COATi, COATi-rev, and PRANK were biologically reasonable for human-gorilla gene pairs, with means of 0.8\%--0.9\% (Fig.~\ref{fig:k2p-empirical}a). Conversely, the means for ClustalΩ, MACSE, and MAFFT were much higher, 1.5\%--1.8\%. These larger means are due to the fact that the right tails of the distributions for ClustalΩ, MACSE, and MAFFT were larger than the other methods. As expected, the alignment methods that ``overaligned'' sequences (produced more mismatches and fewer gaps) also produced higher evolutionary distances on average (Tab.~\ref{tab:gaps-empirical} and Fig.~\ref{fig:k2p-empirical}a). The differences between the distances inferred by COATi and other methods also shows ClustalΩ, MACSE, and MAFFT producing higher distance estimates for a subset of sequences (Fig.~\ref{fig:k2p-empirical}b).
%
COATi's alignments produced significantly lower K2P distances than ClustalΩ, MACSE, MAFFT, and PRANK. Single-tailed Wilcoxon signed-rank tests on matched data produced p-values $< 2.2\times10^{-16}$. Additionally, COATi's K2P distances did not depend on whether human or gorilla was used as the reference sequence (p-value of $0.4212$ for a two-sided test). Additional analysis of the empirical alignments can be found in Supplementary Methods.

@@

In addition to supporting all three phases of gaps, COATi's other advantage over competing aligners in our tests is at least partially due to its default parameters fitting the human-gorilla coding sequences better than the other aligner's default settings. Different data sets may show different results. However, because COATi uses a statistical model with biologically meaningful parameters, users can easily optimize COATi by using parameters estimated by previous studies or estimating parameters themselves \citep[e.g.][]{zhu2022profiling}. This contrasts with other aligners that have limited ability to customize parameters and/or use parameters without biological interpretations.
